---
title: "Ecological niche modelling in R - part 1"
layout: post
---

There are severl ENM/SDM packages in literature, but [dismo](https://cran.r-project.org/web/packages/dismo/index.html) is the very first and very useful package that include a lot of basic functions in ENM. Other packages may help you automatically do a lot works, but dismo is the one that help you understand what is going on underneath process of data manupulation & niche modeling. Personally, I think dismo is a great tool for methodological studies if your study involves parameter manipulations or a lot of repeated workflows (e.g. a lot of species). 


```{r}
# Load dismo library
library(dismo)
```

# 1. occurrences

## 1.1 Load occurrences  
Typically, there are two ways to get occurrences: 1) load a local csv file into R; 2) download observations from online databases, e.g. [GBIF](http://gbif.org) . There is a *gbif()* function can help automatically download species' occurrences without clicking or searching on GBIF website. The parameters provided by *gbif()* give you a lot flexibilities. See this simple example below.
```{r, eval=FALSE}
#simply type in the species' scientific name
 occ <- gbif("Dasypus novemcinctus")
 write.csv(occ,"d:/temp/Dasypus_novemcinctus.csv")
#read.csv("d:/temp/Dasypus_novemcinctus.csv")
```

You may also automatically go over a list of species you are interested.
```{r, eval=FALSE}
species_list <- c("Dasypus kappleri",
                  "Dasypus hybridus",
                  "Dasypus novemcinctus")
for (i in 1:length(species_list)){
   occ <- gbif( species_list[i] )  # download i-th species and save data
   write.csv(occ,paste("d:/temp/",species_list[i],".csv",sep=""))
}
```

Most commonly, we load a csv file from local disk.
```{r}
occ <- read.csv("d:/temp/Dasypus_novemcinctus.csv")
```

## 1.2 Manipulate occurrences  
We always need to clean the occurrences. Here I showed a few functions frequently used for my work.

Example of removing duplicated records according to the columns of lon & lat.
```{r}
# check duplicates
dup_results <- duplicated(occ[c("lon","lat")])
table(dup_results)  # TRUE: not unique record, FALSE: unique record
```
```{r}
nrow(occ)  # original records number
occ_unique <- occ[!dup_results,]
nrow(occ_unique)  # updated records number
```

Example of subseting records by attributes (column).
```{r, message=FALSE, warning=FALSE}
# remove NA values 
occ_updated <- subset(occ_unique, (!is.na(lon)) | (!is.na(lat)) )

# select by year
occ_selected <- subset(occ_unique,year==2016)  # select records in 2016

occ_selected <- subset(occ_unique,year>=1950 & year <=2000)  # between 1950~2000

occ_selected <- subset(occ_unique,year<1950 | year >2000)  # before 1950 or after 2000

occ_selected <- subset(occ_unique,year>=1950  # a combination of more conditions
                       & year <=2000 
                       & lat > 0)


```

## 1.3 Make occurrences spatial!
Pay attention to the grammar.  
(1) Use **<- ~**, not **<-**.  
(2) "lon" and "lat" are column names from the data.frame.  
(3) make sure x/longitude first.  


```{r}
coordinates( occ_updated ) <- ~ lon + lat
class(occ_updated)
```

With R, we can easily transform spatial data into shape files, which may be used in ArcGIS or QGIS for plotting purpose.
```{r,eval=FALSE}
shapefile(occ_updated,"d:/temp/occ.shp")
occ_updated <- shapefile("d:/temp/occ.shp")
```


# 2. Raster files
I used worldclim dataset as an example. Please refer to my  [**GIS Tutorial**](http://lab.fengxiao.info/2015/10/25/GIS-in-R-workshop.html) for more examples of raster analysis in R.  
```{r}
# load two layers: annual T. & P.
basemap <- stack(c("d:/Xiao/GISDATA/climate/WC/WorldClim_10arcmin_bil/bio1.bil"),
                 "d:/Xiao/GISDATA/climate/WC/WorldClim_10arcmin_bil/bio12.bil")

# Extract the raster to North & South America
extent <- c(-180,-30,-60,80)
basemap <- crop(basemap,extent) 
plot(basemap)
```

Plot occurrences on the base map.
```{r,fig.keep = 'last'}
# It is always good to unify the coordinate reference system (CRS). There is a good document about CRS 
# https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf
crs(occ_updated) <- crs(basemap)  # Here I directly assign WGS84 to occurrences; but if you have different CRS, you need to transform one to the other.
plot(basemap[[1]])
plot(occ_updated,add=TRUE,col="red")
```

It seems some points are not located on the landmass, which should have NA values. Let's remove them.
```{r,fig.keep = 'last'}
extracted <- extract(basemap[[1]],occ_updated)
table(is.na(extracted))  # There are a few points have NA values
occ_updated <- occ_updated[!is.na(extracted), ]
plot(basemap[[1]])
plot(occ_updated,add=TRUE,col="red")  # Yes, they disappeared.
```

# 3. Train models in R
## 3.1 Maxent
The following 3 conditions are important for running maxent in R.  
(1) Install **rJava** package.  
(2) I suggest use 32-bit R, because I assume there is an issue of rJava for 64-bit R.   
(3) Also, we must download and copy the maxent.jar into a special folder, typically it is the *java* folder under *dismo* folder. In my case, it is:
**D:/Program Files/R/R-3.3.1/library/dismo/java/maxent.jar**  

There are two ways to run Maxent. One way is similar as the method when we run the interface of maxent, we need occurrence points and rasters; the other way is provide a data.frame with environmental conditions for presences and pseudo-absences.  

Example of the 1st method. Different from the maxent interface, training model, projecting model, and evaluating model are typically independent steps. 

```{r, message=FALSE, warning=FALSE}
# training model
myModel <- maxent(x=basemap,p=occ_updated)

# projecting model 
prediction_now <- predict (myModel, basemap)

# make future temperature warmer & wetter
futuremap <- basemap + 100
prediction_future <- predict (myModel, futuremap)
plot(stack(prediction_now,prediction_future))
```

# More examples to be added:
* seperate occurrences into training and testing
* manipulate training area
* run maxent with data.frame
* manipulate maxent parameters, e.g. beta, features, clamping, output folder


# References:  
* Robert J. Hijmans and Jane Elith (2016) Species distribution modeling with R https://cran.r-project.org/web/packages/dismo/vignettes/sdm.pdf